Using Sweat Na and Cl to Predict CFTR-RD and CRMS

### 2024-03-25 Edith Zemanick/Sweat Test/sweat_test_cf_conference.qmd
cf research conference pretty much final draft

---
title: "Using Sweat Na/Cl Ratio to Predict CFTR-RD and CRMS"
author: "Tim Vigers"
date: "today"
date-format: long
format: revealjs
editor: source
bibliography: Presentation.bib
csl: /Users/timvigers/GitHub/styles/american-medical-association-10th-edition.csl
---

```{r setup}
#| include: false
library(DiagrammeR)
library(readxl)
library(arsenal)
library(tidyverse)
library(factoextra)
library(pROC)
library(knitr)
library(kableExtra)
library(caret)
library(rpart)
library(rpart.plot)
source("/Users/timvigers/GitHub/CF/genotype_class_severity.R")
switch(Sys.info()[["sysname"]],
  Windows = {
    home_dir <- "C:/Users/timvigers/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Edith Zemanick/Sweat Test"
  },
  Darwin = {
    home_dir <- "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Edith Zemanick/Sweat Test"
  }
)
knitr::opts_knit$set(root.dir = home_dir)
```

```{r data cleaning}
# Import exclusion data from Excel
diagnoses <- read_excel("./Data_Cleaned/AM reclassifications All Patients with CF or Non-CF.xlsx")
exclude <- diagnoses$PatientID[!is.na(diagnoses$Exclude) | diagnoses$Code == "Exclude"]
exclude <- exclude[!is.na(exclude)]
# Import data
df <- read.csv("./Data_Raw/Copy of Sweat Tests_AllFor202558_revised01Feb2021.csv", na.strings = "")
n1 <- nrow(df)
n2 <- length(unique(df$PatientID))
# Remove exclusions from Angela
df <- df[!(df$PatientID %in% exclude), ]
# Diagnosis per Angela
df$Diagnosis <- factor(diagnoses$Diagnosis[match(df$PatientID, diagnoses$PatientID)],
  levels = c("CFTR-related disorder", "CRMS", "Cystic Fibrosis", "Non-CF"),
  labels = c("CFTR-RD or CRMS", "CFTR-RD or CRMS", "Cystic Fibrosis", "Non-CF")
)
# Add modulator info
cftr <- read.csv("./Data_Raw/Copy of Report CFTR Modulators.csv")
df$Modulator <- apply(df, 1, function(r) {
  mod <- cftr[cftr$Patient.ID == as.numeric(r["PatientID"]), ]
  if (any(duplicated(mod$Age.at.Start..years.))) {
    mod <- mod[-which(duplicated(mod$Age.at.Start..years.)), ]
  }
  mod <- mod[order(mod$Age.at.Start..years.), ]
  age <- as.numeric(r["Age.at.Test.in.Years"])
  as.character(cut(age, c(-Inf, mod$Age.at.Start..years., Inf),
    labels = c("None", mod$Modulator), right = F
  ))
})
# Numeric variables
num_vars <- c("Amount", "Na", "K", "Cl")
num_vars <- c(paste0(num_vars, "_Left"), paste0(num_vars, "_Right"))
df[, num_vars] <- suppressWarnings(lapply(df[, num_vars], as.numeric))
# Filter per Maxie's code
exclude2 <- c(22, 23, 24, 29, 31, 474, 483, 484, 485, 505, 510, 785, 791, 792, 793)
no_dem <- unique(df$PatientID[df$Demographics.Available != "TRUE" | is.na(df$Demographics.Available) | df$Sex == ""])
df <- df %>%
  mutate(Age.at.Test.in.Days = Age.at.Test.in.Weeks * 7) %>%
  filter(
    Age.at.Test.in.Years < 50 & Age.at.Test.in.Days >= 2,
    !(Diagnosis == "Non-CF" & PatientID %in% exclude2),
    Modulator == "None"
  ) %>%
  distinct(.) %>%
  filter(Sex != "", Demographics.Available == "TRUE")
# Genotype info
## By severity
df$geno1 <- cf_genotype_class_severity(df$Genotypes1)
df$geno2 <- cf_genotype_class_severity(df$Genotypes2)
df$geno1[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno2[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno_risk <- paste(df$geno1, df$geno2, sep = "_")
df$geno_risk <- factor(df$geno_risk,
  levels = c(
    "Severe_Severe", "Mild_Severe", "Mild_Unknown", "Severe_Mild", "Mild_Mild",
    "Not CF_Not CF", "Severe_Unknown", "Unknown_Severe", "Unknown_Unknown"
  ),
  labels = c(
    "High", "Low", "Low", "Low", "Low", "Not CF", "Unknown",
    "Unknown", "Unknown"
  )
)
## By function
df$geno1_class <- cf_genotype_class_severity(df$Genotypes1, out = "class")
df$geno1_class[grep("508", df$Genotypes1)] <- "F508del"
df$geno1_class <- replace(df$geno1_class, df$geno1_class %in% c("I", "II", "III"), "MF")
df$geno1_class <- replace(df$geno1_class, df$geno1_class %in% c("IV", "V"), "RF")

df$geno2_class <- cf_genotype_class_severity(df$Genotypes2, out = "class")
df$geno2_class[grep("508", df$Genotypes2)] <- "F508del"
df$geno2_class <- replace(df$geno2_class, df$geno2_class %in% c("I", "II", "III"), "MF")
df$geno2_class <- replace(df$geno2_class, df$geno2_class %in% c("IV", "V"), "RF")

df$geno_class <- factor(paste0(
  pmin(df$geno1_class, df$geno2_class), "/",
  pmax(df$geno1_class, df$geno2_class)
))
# Which arm?
df$Arm <- apply(df[, c("Cl", "Cl_Left", "Cl_Right", "Na", "Na_Left", "Na_Right", "K", "K_Left", "K_Right")], 1, function(r) {
  k <- sum(!is.na(r))
  if (k == 6) {
    return("Both")
  } else if (k == 3) {
    return("Arm Not Noted")
  } else {
    return(NA)
  }
})
# Merge left and right values
df$Amount <- apply(df[, c("Amount", "Amount_Left", "Amount_Right")], 1, function(r) {
  amt <- suppressWarnings(as.numeric(as.character(r)))
  mean(amt, na.rm = T)
})
df$Cl <- apply(df[, c("Cl", "Cl_Left", "Cl_Right")], 1, function(r) {
  cl <- suppressWarnings(as.numeric(as.character(r)))
  mean(cl, na.rm = T)
})
df$Na <- apply(df[, c("Na", "Na_Left", "Na_Right")], 1, function(r) {
  na <- suppressWarnings(as.numeric(as.character(r)))
  mean(na, na.rm = T)
})
df$K <- apply(df[, c("K", "K_Left", "K_Right")], 1, function(r) {
  k <- suppressWarnings(as.numeric(as.character(r)))
  mean(k, na.rm = T)
})
# Sodium to chloride ratio
df$`Na/Cl Ratio` <- df$Na / df$Cl
high_ratio <- unique(which(df$`Na/Cl Ratio` > 2))
high_cl <- unique(which(df$Cl > 160))
high_na <- unique(which(df$Na > 160))
df <- df[-unique(c(high_ratio, high_cl, high_na)), ]
# Sweat chloride group
df$`Sweat Chloride Group` <- cut(df$Cl, c(-Inf, 30, 60, Inf),
  right = F,
  labels = c("< 30", "30-59", ">60")
)
# Sort
df <- df %>% arrange(PatientID, Age.at.Test.in.Weeks)
# Table options (most variables are not normally distributed)
mycontrols <- tableby.control(
  numeric.test = "kwt",
  numeric.stats = c("Nmiss", "median", "q1q3")
)
labels <- list(
  Age.at.Test.in.Years = "Age (Years)",
  geno_risk = "Genotype Severity", geno_class = "Specific Genotype"
)
# First measure for everyone
t1_df <- df %>%
  group_by(PatientID) %>%
  filter(row_number() == 1)
# Order groups
df$Diagnosis <- factor(df$Diagnosis, levels = c("Cystic Fibrosis", "CFTR-RD or CRMS", "Non-CF"))
```

@@

@@

```{r}
roc_cf <- roc(
  response = df$`Diagnosis`,
  predictor = df$`Na/Cl Ratio`,
  levels = c("CFTR-RD or CRMS", "Cystic Fibrosis"),
  quiet = T
)
coords_cf <- round(coords(roc_cf, "best", best.method = "youden"), 3)
plot(roc_cf, print.auc = TRUE, col = "blue",print.thres=T)
```

@@

```{r}
roc_non_cf <- roc(
  response = df$`Diagnosis`,
  predictor = df$`Na/Cl Ratio`,
  levels = c("Non-CF", "CFTR-RD or CRMS"),
  quiet = T
)
coords_non_cf <- round(coords(roc_non_cf, "best", best.method = "youden"), 3)
plot(roc_non_cf, print.auc = TRUE, col = "red",print.thres=T)
```

@@

@@

@@

```{r}
set.seed(1017)
cart_full <- rpart(Diagnosis ~ Na + Cl, data = cart_df, method = "class",
                   control = rpart.control(cp=0))
rpart.plot(cart_full)
```

@@

```{r}
t <- table(predict(cart_full,type = "class"),cart_df$Diagnosis)
t <- addmargins(t)
# colorize the diagonal elements in table with non-ints
for (i in 1:3) {
  t[i, i] <- cell_spec(t[i, i],
    bold = T, color = "white", background = "darkgreen"
  )
}
kable(t)
```

@@

```{r}
t <- table(predict(cart_full,type = "class"),cart_df$Diagnosis)
t <- round(prop.table(t, margin = 2) * 100, 1)
# colorize the diagonal elements in table with non-ints
for (i in 1:(ncol(t))) {
  t[i, i] <- cell_spec(t[i, i],
    bold = T, color = "white", background = "darkgreen"
  )
}
kable(t, digits = 3)
```

@@

```{r}
set.seed(1017)
cf_crms <- cart_df[cart_df$Diagnosis != "Non-CF", ]
cf_crms$Diagnosis <- droplevels(cf_crms$Diagnosis)
cart_cf_crms <- rpart(Diagnosis ~ Na + Cl, data = cf_crms, method = "class")
rpart.plot(cart_cf_crms)
```

@@

```{r}
t <- table(predict(cart_cf_crms,type = "class"),cf_crms$Diagnosis)
t <- addmargins(t)
# colorize the diagonal elements in table with non-ints
for (i in 1:2) {
  t[i, i] <- cell_spec(t[i, i],
    bold = T, color = "white", background = "darkgreen"
  )
}
kable(t)
```

@@

```{r}
t <- table(predict(cart_cf_crms,type = "class"),cf_crms$Diagnosis)
t <- round(prop.table(t, margin = 2) * 100, 1)
# colorize the diagonal elements in table with non-ints
for (i in 1:(ncol(t))) {
  t[i, i] <- cell_spec(t[i, i],
    bold = T, color = "white", background = "darkgreen"
  )
}
kable(t, digits = 3)
```

@@

```{r}
set.seed(1017)
non_cf_crms <- cart_df[cart_df$Diagnosis != "Cystic Fibrosis", ]
non_cf_crms$Diagnosis <- droplevels(non_cf_crms$Diagnosis)
cart_non_cf_crms <- rpart(Diagnosis ~ Na + Cl, data = non_cf_crms,
                          method = "class")
rpart.plot(cart_non_cf_crms)
```

@@

```{r}
t <- table(predict(cart_non_cf_crms,type = "class"),non_cf_crms$Diagnosis)
t <- addmargins(t)
# colorize the diagonal elements in table with non-ints
for (i in 1:2) {
  t[i, i] <- cell_spec(t[i, i],
    bold = T, color = "white", background = "darkgreen"
  )
}
kable(t)
```

@@

```{r}
t<- table(predict(cart_non_cf_crms,type = "class"),non_cf_crms$Diagnosis)
t <- round(prop.table(t, margin = 2) * 100, 1)
# colorize the diagonal elements in table with non-ints
for (i in 1:(ncol(t))) {
  t[i, i] <- cell_spec(t[i, i],
    bold = T, color = "white", background = "darkgreen"
  )
}
kable(t, digits = 3)
```

# Random forests

@@

@@

  - Support vector machines may allow for more flexeible partitioning of the Na/Cl space.

2. Validate final prediction models on new data.

# Questions?

>>>>

---
title: "Using Sweat Na and Cl to Predict CFTR-RD and CRMS"
author: "Tim Vigers"
date: "today"
date-format: long
format: revealjs
editor: source
bibliography: Presentation.bib
csl: /Users/timvigers/GitHub/styles/american-medical-association-10th-edition.csl
---

```{r setup}
#| include: false
library(DiagrammeR)
library(readxl)
library(arsenal)
library(tidyverse)
library(factoextra)
library(pROC)
library(knitr)
library(kableExtra)
library(caret)
library(rpart)
library(rpart.plot)
library(gtsummary)
source("/Users/timvigers/GitHub/CF/genotype_class_severity.R")
switch(Sys.info()[["sysname"]],
  Windows = {
    home_dir <- "C:/Users/timvigers/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Edith Zemanick/Sweat Test"
  },
  Darwin = {
    home_dir <- "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Edith Zemanick/Sweat Test"
  }
)
knitr::opts_knit$set(root.dir = home_dir)
```

```{r data cleaning}
# Import exclusion data from Excel
diagnoses <- read_excel("./Data_Cleaned/AM reclassifications All Patients with CF or Non-CF.xlsx")
exclude <- diagnoses$PatientID[!is.na(diagnoses$Exclude) |
  diagnoses$Code == "Exclude"]
exclude <- exclude[!is.na(exclude)]
# Import data
df <- read.csv("./Data_Raw/Copy of Sweat Tests_AllFor202558_revised01Feb2021.csv",
  na.strings = ""
)
n1 <- nrow(df)
n2 <- length(unique(df$PatientID))
# Remove exclusions from Angela
df <- df[!(df$PatientID %in% exclude), ]
# Diagnosis per Angela
df$Diagnosis <- factor(diagnoses$Diagnosis[match(df$PatientID, diagnoses$PatientID)],
  levels = c("CFTR-related disorder", "CRMS", "Cystic Fibrosis", "Non-CF"),
  labels = c("CFTR-RD or CRMS", "CFTR-RD or CRMS", "Cystic Fibrosis", "Non-CF")
)
# Add modulator info
cftr <- read.csv("./Data_Raw/Copy of Report CFTR Modulators.csv")
df$Modulator <- apply(df, 1, function(r) {
  mod <- cftr[cftr$Patient.ID == as.numeric(r["PatientID"]), ]
  if (any(duplicated(mod$Age.at.Start..years.))) {
    mod <- mod[-which(duplicated(mod$Age.at.Start..years.)), ]
  }
  mod <- mod[order(mod$Age.at.Start..years.), ]
  age <- as.numeric(r["Age.at.Test.in.Years"])
  as.character(cut(age, c(-Inf, mod$Age.at.Start..years., Inf),
    labels = c("None", mod$Modulator), right = F
  ))
})
# Numeric variables
num_vars <- c("Amount", "Na", "K", "Cl")
num_vars <- c(paste0(num_vars, "_Left"), paste0(num_vars, "_Right"))
df[, num_vars] <- suppressWarnings(lapply(df[, num_vars], as.numeric))
# Filter per Maxie's code
exclude2 <- c(
  22, 23, 24, 29, 31, 474, 483, 484, 485, 505, 510, 785, 791, 792, 793
)
no_dem <- unique(df$PatientID[df$Demographics.Available != "TRUE" | is.na(df$Demographics.Available) | df$Sex == ""])
df <- df %>%
  mutate(Age.at.Test.in.Days = Age.at.Test.in.Weeks * 7) %>%
  filter(
    Age.at.Test.in.Years < 50 & Age.at.Test.in.Days >= 2,
    !(Diagnosis == "Non-CF" & PatientID %in% exclude2),
    Modulator == "None"
  ) %>%
  distinct(.) %>%
  filter(Sex != "", Demographics.Available == "TRUE")
# Genotype info
## By severity
df$geno1 <- cf_genotype_class_severity(df$Genotypes1)
df$geno2 <- cf_genotype_class_severity(df$Genotypes2)
df$geno1[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno2[df$Diagnosis == "Non-CF"] <- "Not CF"
df$geno_risk <- paste(df$geno1, df$geno2, sep = "_")
df$geno_risk <- factor(df$geno_risk,
  levels = c(
    "Severe_Severe", "Mild_Severe", "Mild_Unknown", "Severe_Mild", "Mild_Mild",
    "Not CF_Not CF", "Severe_Unknown", "Unknown_Severe", "Unknown_Unknown"
  ),
  labels = c(
    "High", "Low", "Low", "Low", "Low", "Not CF", "Unknown",
    "Unknown", "Unknown"
  )
)
## By function
df$geno1_class <- cf_genotype_class_severity(df$Genotypes1, out = "class")
df$geno1_class[grep("508", df$Genotypes1)] <- "F508del"
df$geno1_class <- replace(
  df$geno1_class,
  df$geno1_class %in% c("I", "II", "III"), "Minimal function"
)
df$geno1_class <- replace(
  df$geno1_class,
  df$geno1_class %in% c("IV", "V"), "Residual function"
)

df$geno2_class <- cf_genotype_class_severity(df$Genotypes2, out = "class")
df$geno2_class[grep("508", df$Genotypes2)] <- "F508del"
df$geno2_class <- replace(
  df$geno2_class,
  df$geno2_class %in% c("I", "II", "III"), "Minimal function"
)
df$geno2_class <- replace(
  df$geno2_class,
  df$geno2_class %in% c("IV", "V"), "Residual function"
)

df$geno_class <- factor(paste0(
  pmin(df$geno1_class, df$geno2_class), "/",
  pmax(df$geno1_class, df$geno2_class)
))
# Which arm?
df$Arm <- apply(df[, c(
  "Cl", "Cl_Left", "Cl_Right", "Na", "Na_Left", "Na_Right", "K", "K_Left",
  "K_Right"
)], 1, function(r) {
  k <- sum(!is.na(r))
  if (k == 6) {
    return("Both")
  } else if (k == 3) {
    return("Arm Not Noted")
  } else {
    return(NA)
  }
})
# Merge left and right values
df$Amount <- apply(
  df[, c("Amount", "Amount_Left", "Amount_Right")], 1,
  function(r) {
    amt <- suppressWarnings(as.numeric(as.character(r)))
    mean(amt, na.rm = T)
  }
)
df$Cl <- apply(df[, c("Cl", "Cl_Left", "Cl_Right")], 1, function(r) {
  cl <- suppressWarnings(as.numeric(as.character(r)))
  mean(cl, na.rm = T)
})
df$Na <- apply(df[, c("Na", "Na_Left", "Na_Right")], 1, function(r) {
  na <- suppressWarnings(as.numeric(as.character(r)))
  mean(na, na.rm = T)
})
df$K <- apply(df[, c("K", "K_Left", "K_Right")], 1, function(r) {
  k <- suppressWarnings(as.numeric(as.character(r)))
  mean(k, na.rm = T)
})
# Sodium to chloride ratio
df$`Na/Cl Ratio` <- df$Na / df$Cl
high_ratio <- unique(which(df$`Na/Cl Ratio` > 2))
high_cl <- unique(which(df$Cl > 160))
high_na <- unique(which(df$Na > 160))
df <- df[-unique(c(high_ratio, high_cl, high_na)), ]
# Sweat chloride group
df$`Sweat Chloride Group` <- cut(df$Cl, c(-Inf, 30, 60, Inf),
  right = F,
  labels = c("< 30", "30-59", ">60")
)
# Sort
df <- df %>% arrange(PatientID, Age.at.Test.in.Weeks)
# Table options (most variables are not normally distributed)
mycontrols <- tableby.control(
  numeric.test = "kwt",
  numeric.stats = c("Nmiss", "median", "q1q3")
)
labels <- list(
  Age.at.Test.in.Years = "Age (Years)",
  geno_risk = "Genotype Severity", geno_class = "Specific Genotype"
)
# First measure for everyone
t1_df <- df %>%
  group_by(PatientID) %>%
  filter(row_number() == 1)
# Order groups
df$Diagnosis <- factor(df$Diagnosis,
  levels = c("Cystic Fibrosis", "CFTR-RD or CRMS", "Non-CF")
)
```

@@

## Table 1 {.scrollable}

```{r}
df %>%
  filter(!is.na(Na) & !is.na(Cl)) %>%
  select(Diagnosis, Age.at.Test.in.Weeks, Sex, Na, Cl, K, Method, geno_class) %>%
  tbl_summary(
    by = Diagnosis, missing_text = "Missing",
    label = list(
      Age.at.Test.in.Weeks = "Age (Weeks)",
      geno_class = "Genotype Class"
    )
  ) %>%
  add_overall() %>%
  add_p(test = list(geno_class ~ "fisher.test"), test.args = all_tests("fisher.test") ~ list(simulate.p.value = TRUE)) %>%
  separate_p_footnotes()
```

@@

```{r}
roc_cf <- roc(
  response = df$`Diagnosis`,
  predictor = df$`Na/Cl Ratio`,
  levels = c("CFTR-RD or CRMS", "Cystic Fibrosis"),
  quiet = T
)
coords_cf <- round(coords(roc_cf, "best", best.method = "youden"), 3)
plot(roc_cf, print.auc = TRUE, col = "blue", print.thres = T)
```

@@

```{r}
roc_non_cf <- roc(
  response = df$`Diagnosis`,
  predictor = df$`Na/Cl Ratio`,
  levels = c("Non-CF", "CFTR-RD or CRMS"),
  quiet = T
)
coords_non_cf <- round(coords(roc_non_cf, "best", best.method = "youden"), 3)
plot(roc_non_cf, print.auc = TRUE, col = "red", print.thres = T)
```

@@

## K-means pros and cons

| Pros                                         | Cons                                                        |
|----------------------------------------------|-------------------------------------------------------------|
| Easy to implement and understand.            | Must choose k manually.                                     |
| Scales well to large data sets.              | Has trouble when clusters are of varying sizes and density.                                                            |
| Easily adapts to new examples.               | Centroids can be dragged by outliers. |
| Generalizes to clusters of different shapes. | Does not scale well with more dimensions.                        |

@@

## Classification tree pros and cons

| Pros                                      | Cons                                                                |
|-------------------------------------------|---------------------------------------------------------------------|
| Easy to interpret.                        | Prone to overfitting.                                               |
| Robust to outliers (except when overfit). | Unstable unless well- regularized.                                  |
| Non-linear.                               | Greedy algorithm means the tree is not and optimal global solution. |
| Non-parametric.                           | Can become unwieldy if not pruned correctly.                        |

@@

```{r}
set.seed(1017)
cart_full <- rpart(Diagnosis ~ Na + Cl,
  data = cart_df, method = "class",
  control = rpart.control(cp = 0)
)
rpart.plot(cart_full, type = 3, extra = 8)
```

@@

```{r}
t <- table(predict(cart_full, type = "class"), cart_df$Diagnosis)
t <- addmargins(t)
# colorize the diagonal elements in table with non-ints
for (i in 1:3) {
  t[i, i] <- cell_spec(t[i, i],
    bold = T, color = "white", background = "darkgreen"
  )
}
kable(t)
```

@@

```{r}
t <- table(predict(cart_full, type = "class"), cart_df$Diagnosis)
t <- round(prop.table(t, margin = 2) * 100, 1)
# colorize the diagonal elements in table with non-ints
for (i in 1:(ncol(t))) {
  t[i, i] <- cell_spec(t[i, i],
    bold = T, color = "white", background = "darkgreen"
  )
}
kable(t, digits = 3)
```

@@

```{r}
set.seed(1017)
cf_crms <- cart_df[cart_df$Diagnosis != "Non-CF", ]
cf_crms$Diagnosis <- droplevels(cf_crms$Diagnosis)
cart_cf_crms <- rpart(Diagnosis ~ Na + Cl, data = cf_crms, method = "class")
rpart.plot(cart_cf_crms, type = 3, extra = 8)
```

@@

```{r}
t <- table(predict(cart_cf_crms, type = "class"), cf_crms$Diagnosis)
t <- addmargins(t)
# colorize the diagonal elements in table with non-ints
for (i in 1:2) {
  t[i, i] <- cell_spec(t[i, i],
    bold = T, color = "white", background = "darkgreen"
  )
}
kable(t)
```

@@

```{r}
t <- table(predict(cart_cf_crms, type = "class"), cf_crms$Diagnosis)
t <- round(prop.table(t, margin = 2) * 100, 1)
# colorize the diagonal elements in table with non-ints
for (i in 1:(ncol(t))) {
  t[i, i] <- cell_spec(t[i, i],
    bold = T, color = "white", background = "darkgreen"
  )
}
kable(t, digits = 3)
```

@@

```{r}
set.seed(1017)
non_cf_crms <- cart_df[cart_df$Diagnosis != "Cystic Fibrosis", ]
non_cf_crms$Diagnosis <- droplevels(non_cf_crms$Diagnosis)
cart_non_cf_crms <- rpart(Diagnosis ~ Na + Cl,
  data = non_cf_crms,
  method = "class"
)
rpart.plot(cart_non_cf_crms, type = 1, extra = 8)
```

@@

```{r}
t <- table(predict(cart_non_cf_crms, type = "class"), non_cf_crms$Diagnosis)
t <- addmargins(t)
# colorize the diagonal elements in table with non-ints
for (i in 1:2) {
  t[i, i] <- cell_spec(t[i, i],
    bold = T, color = "white", background = "darkgreen"
  )
}
kable(t)
```

@@

```{r}
t <- table(predict(cart_non_cf_crms, type = "class"), non_cf_crms$Diagnosis)
t <- round(prop.table(t, margin = 2) * 100, 1)
# colorize the diagonal elements in table with non-ints
for (i in 1:(ncol(t))) {
  t[i, i] <- cell_spec(t[i, i],
    bold = T, color = "white", background = "darkgreen"
  )
}
kable(t, digits = 3)
```

# Support vector machines

@@

- Aims to find a hyperplane that best separates our groups.
  - In a p-dimensional space, a hyperplane is a flat affine subspace of dimension p − 1.
    - When considering 2 dimensions, this is just a line.

## Hyperplane classifiers

![](hyperplane.png)

## The maximal margin classifier

![](mmc.png)

The points with arrows to our dividing line are the *support vectors*.

## But what if the boundary isn't just a simple line?

![](nlb.png)

## But what if the boundary isn't just a simple line?

- The *support vector machine* (SVM) is an extension of the support vector classifier where the feature space is enlarged using kernels.
  - This is a little complicated, but is analogous to adding polynomial terms to a regression model.
    - In fact, SVMs are closely related to logistic regression.

## SVM pros and cons

- One of the best “out of the box” classifiers.
- Works best when there's fairly clear separation between classes.

## A simple SVM with radial kernel

![](plot_zoom.png)

```{r}
library(e1071)
set.seed(1017)
svmfit <- svm(Diagnosis ~ Na + Cl, data = cart_df)
```

```{r}
t <- table(predict(svmfit, type = "class"), cart_df$Diagnosis)
t <- addmargins(t)
# colorize the diagonal elements in table with non-ints
for (i in 1:2) {
  t[i, i] <- cell_spec(t[i, i],
    bold = T, color = "white", background = "darkgreen"
  )
}
kable(t)
```

```{r}
t <- table(predict(svmfit, type = "class"), cart_df$Diagnosis)
t <- round(prop.table(t, margin = 2) * 100, 1)
# colorize the diagonal elements in table with non-ints
for (i in 1:(ncol(t))) {
  t[i, i] <- cell_spec(t[i, i],
    bold = T, color = "white", background = "darkgreen"
  )
}
kable(t, digits = 3)
```

## Comparison of methods

| Method                                           | % Correct CF | % Correct Non-CF | % Correct CFTR-RD or CRMS |
|--------------------------------------------------|--------------|------------------|---------------------------|
| K-means                                          | 93.1         | 81               | 78.6                      |
| Classification tree (all)                        | 98.9         | 99.9             | 21.4                      |
| Classification tree (CF vs. CFTR-RD or CRMS)     | 99.2         |                  | 92.9                      |
| Classification tree (Non-CF vs. CFTR-RD or CRMS) | 39.3         | 99.5             |                           |
| SVM                                              | 98.2         | 99.9             | 0                         |

@@

  - Bagging, boosting, and random forests are all extensions of regression trees that minimize overfitting.

2. Scrap SVMs for now.

3. Would it be worth considering additional variables for these models? Sex, age at test, etc.?

4. Validate final prediction models on new data.

# Questions, comments, suggestions?
